Circular Distributional Trees and Forests with an Application to Wind Direction Forecasting

### 2019-11-22 pkg/circtree/inst/wind_paper/draft/draft.Rnw
minor revisions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2895 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: e4133dbdd6289ad239ad1444e5b20980384adf54

In many approaches to model circular data it is assumed that the circular
variable of interest follows a circular distribution, in particular the von
Mises distribution which is also known as the ``circular normal distribution''
(see Section~\ref{sec:prob_circ} for more details). This is also the basis
which early regression models for circular data are built on as they regress
the parameter(s) of the von Mises distribution on covariates. One of the first
regression models with a circular response variable and linear covariates was
presented by \cite{Gould:1969} where the circular mean, i.e., the location
parameter of the von Mises distribution is predicted by a linear combination of
covariates. \cite{Johnson+Wehrly:1978} took this idea one step further by
plugging in a link function such that the values of the linear predictor are
transformed to a restricted interval of length $2\pi$. This generalized linear
model (GLM)-type approach was further extended by \cite{Fisher+Lee:1992} and
subsequently by \cite{Fisher:1993} introducing another specific transformation
function which maps the values of the real line to the unit circle. Moreover,
the presented methodology allows for both distribution parameters, the mean
direction and the concentration parameter of the von Mises distribution, to be
modeled depending on covariates, not only in separate models but also in a
mixed model. While being built on well elaborated theory this methodology has
shown to be very challenging in their application, mainly due to the complexity
encountered in optimizing the log-likelihood function which is not globally
concave. Therefore, highly informative starting values are crucial for the
model to converge at all \citep{Pewsey+Neuhaeuser+Ruxton:2013,
Gill+Hangartner:2010}. In order to avoid this strong dependence on appropriate
initial values, \cite{Mulder+Klugkist:2017} presented an alternative approach
to regression modeling of circular data by employing a GLM in a Bayesian
framework. The listed methods and most of their further developments
\citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} are built on additive
regression models. While these generally provide a powerful toolbox for
modeling particularly smooth effects, they may not be able to capture all types
of effects adequately. Especially abrupt changes might not be detected
properly owing to the enforced additive structure and model assumptions, such
as a possibly highly non-linear link function and the form of linear
combinations.

@@

% -------------------------------------------------------------------
% Probabilistic circular modeling
% -------------------------------------------------------------------
\section{Probabilistic circular modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a proper
circular probability distribution, i.e., a probability distribution which
accounts for the periodicity of circular data. Generally, this feature can be
obtained by ``wrapping'' the probability density function of any continuous
distribution around the unit circle \citep{Mardia+Jupp:2009}. In that way, the
wrapped Cauchy distribution or the wrapped normal distribution can be employed
to model symmetric unimodal circular data. A close approximation to the wrapped
normal distribution that is mathematically simpler and hence easier to use
\citep{Fisher:1993} is provided by the von Mises distribution, a purely
circular distribution which is also known as ``the circular normal
distribution'' and commonly applied for probabilistic modeling of circular
data. Based on a location parameter $\mu \in (-\pi, \pi]$ and a concentration
parameter $\kappa > 0$ the density of the von Mises distribution for an
observation $y \in (-\pi, \pi]$ is given by:
%%
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
\end{equation}
%%
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order
$0$ \citep[see, e.g.,][for a more detailed
overview]{Jammalamadaka+Sengupta:2001}.

@@

By means of the score function
%%
\begin{equation}
  \label{eq:scores}
  \begin{aligned}
    s(\mu, \kappa, y) &= \bigg(\frac{\partial \ell}{\partial \mu}(\mu,\kappa; y)
    \smallskip &,
    \quad & \frac{\partial \ell}{\partial \kappa}(\mu,\kappa; y_i) \bigg) 
    \\
    &= \bigg(\kappa \sin(y-\mu)
    \smallskip &,
    \quad &-\frac{I_1(\kappa)}{2\pi I_0(\kappa)}+\cos(y-\mu) \bigg)
  \end{aligned}
\end{equation}
%%
an indicator of deviation and in that way a measure of goodness of fit of the
model can be obtained for each observation $y_i$ and fitted parameter.
Moreover, the optimization problem in Equation~\ref{eq:optim} can also be
defined by 
%%
\begin{equation}
  \sum_{i = 1}^n s(\hat{\mu}, \hat{\kappa}, y_i) = 0.
\end{equation}
%%
%%
\begin{figure}[t]
  \begin{minipage}{.05\textwidth}
    \hfill
  \end{minipage}%
  \begin{minipage}{.45\textwidth}
    \vspace{-1em}
    \includegraphics[width=1.05\linewidth]{density_linear.pdf}
  \end{minipage}%
  \begin{minipage}{.45\textwidth}
    \includegraphics[width=1.2\linewidth]{density_circular.pdf}
  \end{minipage}
  \begin{minipage}{.05\textwidth}
    \hfill
  \end{minipage}%
  \caption{Illustration of a distributional model fitted to circular data by
    maximum likelihood estimation employing the von Mises distribution. Left
    panel: ``linear'' illustration restricted to the interval $[0, 2\pi)$. Right
    panel: circular illustration. In both panels the empirical histogram (gray)
    and fitted density (red line) are depicted along with the estimated location
    parameter (red hand).}
  \label{fig:densities} 
\end{figure}
%%

@@

\subsection{Circular trees}
\label{sec:circtree}
Fitting a global model to a full data set can be very challenging, particularly
for complex data with strong variations. Therefore, separating the data set
into more homogeneous subgroups based on covariates before fitting a local
model in each of these subgroups allows to capture (possibly) group specific
effects more precisely and hence can result in a overall better fitting model.
This is the general idea of regression trees which are combined with
distributional modeling in \cite{Schlosser+Hothorn+Stauffer:2019} based on the
unbiased recursive partitioning algorithms MOB
\citep{Zeileis+Hothorn+Hornik:2008} or CTree
\citep{Hothorn+Hornik+Zeileis:2006}. By applying the presented distributional
trees, a full distributional model is specified in each node of the tree for a
specific distribution family chosen in advance. Hence, selecting the von Mises
distribution allows for an application of distributional trees to circular data
following the tree building algorithm as introduced in Section~2.2. of
\cite{Schlosser+Hothorn+Stauffer:2019}.

In the applied tree-building algorithm, in each node the crucial decision on
how and where to split the data is based on model scores which are obtained by
evaluating the score function $s$ for the individual observations and parameter
estimates (Equation~\ref{eq:scores}). For the von Mises distribution with its
two distribution parameters and a data set of $n$ observations this yields an
$n \times 2$ matrix that can be employed as a discrepancy measure, capturing
how well each given observation conforms with the estimated location
$\hat{\mu}$ and precision $\hat{\kappa}$, respectively. To capture dependence
on covariates, the association between the model's scores and each available
covariate is assessed using either a parameter instability test~(MOB) or a
permutation test~(CTree). By doing so in each partitioning step, the covariate
with the highest significant association (i.e., lowest significant $p$-value,
if any) is selected for splitting the data. The corresponding split point is
chosen either by optimizing the log-likelihood~(MOB) or a two-sample test
statistic~(CTree) over all possible partitions. This procedure is repeated
recursively until there are no significant parameter instabilities or until
another stopping criterion is met (e.g., subgroup size or tree depth).

@@

\subsection{Data}
\label{sec:wind:data}
The circular response variable considered in this case study is a $10$\,min
mean of wind direction measurements at the airports Innsbruck and Vienna on an
hourly temporal resolution. As predictor variables, we use $1$-hourly resolved
$10$\,min mean observations of various meteorological quantities such as wind
direction, wind speed, temperature, air pressure and humidity, all lagged by
one or three hours according to the respective forecasting step. The
meteorological variables are measured either directly at the airports or within
the their vicinity. For Innsbruck, measurements at the airport and along the
intersecting valleys are used, whereas, for Vienna, measurements at the airport
and within an area of about $10$\,km$^2$ are used. A topographical overview of
the airports and their surrounding areas with the station sites employed in
this study is provided in Figure~XXX. In addition, we use derived quantities
such as 3-hourly means, minima and maxima, as well as 1-and 3-hourly temporal
changes and spatial differences towards the airport of the respective
quantities. An overview of the employed data sets can be found in
Table~\ref{tab:data}.

The data used in this study consists of 5~total years from January 2015 to
December 2018. After first eliminating predictor variables with more than 5\%
missing values and then time points with missing observations, the data set
consists of 41979 time points and 260 covariates for Innsbruck, and of 38985
times points and 494 covariates for Vienna, respectively.

%%
\begin{table}[t!]
  \caption[Table caption text]{Overview of the employed data sets for the case
    study: For Innsbruck and Vienna, various meteorological variables and derived
    quantities of these are considered at the respective stations, located either
    directly at the airports or in their vicinities.}
  \label{tab:data}
  \begin{center}
    \begin{tabular}{l  l}
      \hline
      Data components               & Description \\
      \hline
      Meteorological variables:     & Wind direction, wind (gust) speed, \\
                                    & (reduced) air pressure, relative humidity, \\
                                    & temperature\\ 
      \noalign{\vskip 1mm}                           
      Derived quantities:           & 3-hourly means/minima/maxima, \\
                                    & 1-hourly and 3-hourly temporal changes, \\
                                    & spatial differences towards the airport\\
      \noalign{\vskip 1mm}                              
      Weather stations (Innsbruck): & 4 stations at the airport, \\
                                    & Igls, Kematen, Kufstein, Landeck, Patscherkofel, \\
                                    & Steinach\\
      \noalign{\vskip 1mm}
      Weather stations (Vienna):    & 9 stations at the airport, \\
                                    & Arsenal, Donaufeld, Exelberg, G\"anserndorf, \\
                                    & Gro{\ss}-Enzersdorf, Gumpoldskirchen, Hohe Warte,\\
                                    & Innere Stadt, Jubil\"aumswarte, Mariabrunn, \\
                                    & Seibersdorf, Unterlaa, Wolkersdorf\\
      \hline
    \end{tabular}
  \end{center}
\end{table}
%%

@@

    wide range of applications and are often also used as a baseline for the

@@

    Section~\ref{sec:introduction}, circular regression models can experience the

@@

    appropriate link function~$g$ in order to guarantee response values on the
    interval $(-\pi,\pi]$. As the implementation cannot handle circular covariates,

@@

  \item \emph{Circular distribution forest:} Following the description in

@@

\subsection{Results} 
This section provides a detailed analysis on the predictive performance of the
different proposed statistical models applied to probabilistic wind direction
forecasting. To ensure a comprehensive comparison of the models, wind direction
forecasts are evaluated for two different lead times at two airports with
different climatological site characteristics.
Figure~\ref{fig:boxplot_crpsraw} shows the CRPS values of the employed models
at forecast steps 1\,h and 3\,h for the airports Innsbruck (Panels~a,\,c) and
Vienna (Panels~b,\,d). The scores are aggregated over the $5$~validation years,
yielding yearly mean values for every hour per calendar month, with a lower
score indicating better performance. For both stations at both forecast steps,
the circular forest overall provides the best predictive performance, followed
by the circular distributional tree and, except for the $3$\,h forecast at
Innsbruck, the persistency model. In comparison to the circular tree and
forest, for both stations and forecast steps, the climatological model and the
linear model show clearly higher CRPS values and hence a lower predictive
performance.

The different site characteristics of the airports Innsbruck
(Figure~\ref{fig:boxplot_crpsraw}\,a,\,c) and Vienna
(Figure~\ref{fig:boxplot_crpsraw}\,b,\,d) seem to have an effect on the
absolute level of the model performances and on their respective predictive
performance variances. At Innsbruck, due to the surrounding mountains a
limited number of possible wind directions exists, namely the three wind
regimes discussed for Figure~\ref{fig:tree_ibk} in
Section~\ref{sec:introduction}. Therefore, for Innsbruck the wind direction
remains rather constant in one of these possible states, but once a change
takes place it is mostly a major wind direction shift. Due to the few wind
regimes the rather inflexible climatological and linear models score relatively
well with similar CRPS values as the other models
(Figure~\ref{fig:boxplot_crpsraw}\,a,\,c). In addition, at Innsbruck the
potential high prediction errors in case of a paradigm shift seem to lead to a
higher variation in the predictive performance for all models in comparison to
Vienna; this variation is especially high for the persistency model due to its
strong vulnerability to abrupt wind shifts. On the contrary, at Vienna the
geographic surroundings do not constrain any wind direction which leads to
smaller and less abrupt changes in the wind direction as well as less
pronounced wind regimes. This seems to weaken the predictive performance of the
climatological and linear models, and to reduce the performance variability for
all models (Figure~\ref{fig:boxplot_crpsraw}\,b,\,d).

The different forecast steps have apparently only a minor effect on the
predictive performance of the climatological model and the linear model at both
stations. However, for the persistency model, at both stations, higher scores
for the $3$\,h forecast (Figure~\ref{fig:boxplot_crpsraw}\,c,\,d) reveal a
lower performance for longer lead times; this is probably due to minimum
3-hourly instead of 1-hourly lagged response values employed as covariates in
the persistency model. The circular distributional tree and forest seem to
partially compensate for the lower skill of the lagged response values by other
covariates, hence their predictive performance only slightly decreases for the
longer lead time. This compensation is especially evident for Innsbruck, where
the performance difference between the persistency model and the tree-based
methods significantly increases from the 1-hourly to the 3-hourly forecast.
%%
\begin{figure}[t]
  \centering
  \setkeys{Gin}{width=1.0\textwidth}
  \includegraphics{_plot_circforest_validation_crpsraw_agg_comparison_with_lowff_v14b.pdf}
  \caption{CRPS skill scores for wind direction forecasts based on the full
    predictive von~Mises distribution for $+1$ and $+3$\,h forecasts at the
    airports Innsbruck and Vienna. Each box-and-whisker contains $24$~hourly scores
    per for each of the $12$~months averaged over the $5$~validation years which
    yields a total of $288$~yearly mean values. The scores are shown for the
    climatological, persistency and the linear model as well as for the circular
    distributional tree and forest.}
  \label{fig:boxplot_crpsraw} 
\end{figure}
%%

In addition to the raw CRPS (Figure~\ref{fig:boxplot_crpsraw}), CRPS skill
scores with the climatological model as a reference are provided in
Figure~\ref{fig:boxplot_crpsskill}. Skill scores are in percent, and positive
values indicate an improvement in the predictive performance over the
reference. For all setups, the circular forest has the highest skill scores
with a mean performance gain of 13--25\% and 58--71\% for Innsbruck and Vienna,
respectively. As discussed for Figure~\ref{fig:boxplot_crpsraw}, this
improvement over the climatological model is lower for Innsbruck due to the low
number of predominant wind regimes and hence a relatively good performance of
the climatological model. Additionally, Figure~\ref{fig:boxplot_crpsskill}
shows that even if the persistency model's performance is lower than the
reference (Panel~c) the tree-based models are significantly superior to the
climatological model used as reference.
%%
\begin{figure}[t]
  \centering
  \setkeys{Gin}{width=1.0\textwidth}
  \includegraphics{_plot_circforest_validation_crpsskill_agg_comparison_with_lowff_v14b.pdf}
  \caption{As Figure~\ref{fig:boxplot_crpsraw}, but showing CRPS skill scores
    with the climatology model as reference. Skill scores are in percent; positive
    values indicate improvements over the reference.} 
  \label{fig:boxplot_crpsskill} 
\end{figure}
%%

% -------------------------------------------------------------------
% Summary
% -------------------------------------------------------------------
\section{Summary and conclusion}
\label{sec:summary}
Circular distributional trees for circular responses are established by
coupling model-based recursive partitioning with the von Mises distribution.
The resulting trees can capture non-linear changes, shifts, and potential
interactions in covariates without prespecification of such effects. This is
particularly useful for modeling a highly fluctating response such as wind
direction. \dots

@@

\section*{Computational details}
The proposed circular tree and forest models are implemented in the \textsf{R}
package \textbf{circtree} (version~\Sexpr{packageVersion("circtree")}). The
package is based on the \textbf{disttree} package
(version~\Sexpr{packageVersion("disttree")}) which applies the main tree
building functions from the \textbf{partykit} package
(version~\Sexpr{packageVersion("partykit")}). All three packages are available
on \textsf{R}-Forge at \url{https://R-Forge.R-project.org/projects/partykit/}.

>>>>

In many approaches to model circular data it is assumed that the circular
variable of interest follows a circular distribution, in particular the von
Mises distribution which is also known as the ``circular normal distribution''
(see Section~\ref{sec:prob_circ} for more details). This is also the basis
which early regression models for circular data are built on as they regress
the parameter(s) of the von Mises distribution on covariates. One of the first
regression models with a circular response variable and linear covariates was
presented by \cite{Gould:1969} where the circular mean, i.e., the location
parameter of the von Mises distribution is predicted by a linear combination of
covariates. \cite{Johnson+Wehrly:1978} took this idea one step further by
plugging in a link function such that the values of the linear predictor are
transformed to a restricted interval of length $2\pi$. This generalized linear
model (GLM)-type approach was further extended by \cite{Fisher+Lee:1992} and
subsequently by \cite{Fisher:1993} introducing another specific transformation
function which maps the values of the real line to the unit circle. Moreover,
the presented methodology allows for both distribution parameters, the mean
direction and the concentration parameter of the von Mises distribution, to be
modeled depending on covariates, not only in separate models but also in a
mixed model. While being built on well elaborated theory this methodology has
shown to be very challenging in its application, mainly due to the complexity
encountered in optimizing the log-likelihood function which is not globally
concave. Therefore, highly informative starting values are crucial for the
model to converge at all \citep{Pewsey+Neuhaeuser+Ruxton:2013,
Gill+Hangartner:2010}. In order to avoid this strong dependence on appropriate
initial values, \cite{Mulder+Klugkist:2017} presented an alternative approach
to regression modeling of circular data by employing a GLM in a Bayesian
framework. All listed methods and most of their further developments
\citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} are built on additive
regression models. While these generally provide a powerful toolbox for
modeling particularly smooth effects, they may not be able to capture all types
of effects adequately. Especially abrupt changes might not be detected
properly owing to the enforced additive structure and model assumptions, such
as a possibly highly non-linear link function and the form of linear
combinations.

@@

% -------------------------------------------------------------------
% Probabilistic circular modeling
% -------------------------------------------------------------------
\newpage
\section{Probabilistic circular modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a proper
circular probability distribution, i.e., a probability distribution which
accounts for the periodicity of circular data. Generally, this feature can be
obtained by ``wrapping'' the probability density function of any continuous
distribution around the unit circle \citep{Mardia+Jupp:2009}. In that way, the
wrapped Cauchy distribution or the wrapped normal distribution can be employed
to model symmetric unimodal circular data. A close approximation to the wrapped
normal distribution that is mathematically simpler and hence easier to use
\citep{Fisher:1993} is provided by the von Mises distribution, a purely
circular distribution which is also known as ``the circular normal
distribution'' and commonly applied for probabilistic modeling of circular
data. Based on a location parameter $\mu \in (-\pi, \pi]$ and a concentration
parameter $\kappa > 0$ the density of the von Mises distribution for an
observation $y \in (-\pi, \pi]$ is given by:
%%
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
\end{equation}
%%
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order
$0$ \citep[see, e.g.,][for a more detailed
overview]{Jammalamadaka+Sengupta:2001}.

@@

By means of the score function
%%
\begin{equation}
  \label{eq:scores}
  \begin{aligned}
    s(\mu, \kappa, y) &= \bigg(\frac{\partial \ell}{\partial \mu}(\mu,\kappa; y)
    \smallskip &,
    \quad & \frac{\partial \ell}{\partial \kappa}(\mu,\kappa; y_i) \bigg) 
    \\
    &= \bigg(\kappa \sin(y-\mu)
    \smallskip &,
    \quad &-\frac{I_1(\kappa)}{2\pi I_0(\kappa)}+\cos(y-\mu) \bigg)
  \end{aligned}
\end{equation}
%%
an indicator of deviation and in that way a measure of goodness of fit of the
model can be obtained for each observation and fitted parameter.
Moreover, the optimization problem in Equation~\ref{eq:optim} can also be
defined by 
%%
\begin{equation}
  \sum_{i = 1}^n s(\hat{\mu}, \hat{\kappa}, y_i) = 0.
\end{equation}
%%
%%
\begin{figure}[t]
  \begin{minipage}{.05\textwidth}
    \hfill
  \end{minipage}%
  \begin{minipage}{.45\textwidth}
    \vspace{-1em}
    \includegraphics[width=1.05\linewidth]{density_linear.pdf}
  \end{minipage}%
  \begin{minipage}{.45\textwidth}
    \includegraphics[width=1.2\linewidth]{density_circular.pdf}
  \end{minipage}
  \begin{minipage}{.05\textwidth}
    \hfill
  \end{minipage}%
  \caption{Illustration of a distributional model fitted to circular data by
    maximum likelihood estimation employing the von Mises distribution. Left
    panel: ``linear'' illustration restricted to the interval $[0, 2\pi)$. Right
    panel: circular illustration. In both panels the empirical histogram (gray)
    and fitted density (red line) are depicted along with the estimated location
    parameter (red hand).}
  \label{fig:densities} 
\end{figure}
%%

@@

\subsection{Circular trees}
\label{sec:circtree}
Fitting a global model to a full data set can be very challenging, particularly
for complex data with strong variations. Therefore, separating the data set
into more homogeneous subgroups based on covariates before fitting a local
model in each of these subgroups allows to capture (possibly) group specific
effects more precisely and hence can result in an overall better fitting model.
This is the general idea of regression trees which are combined with
distributional modeling in \cite{Schlosser+Hothorn+Stauffer:2019} based on the
unbiased recursive partitioning algorithms MOB
\citep{Zeileis+Hothorn+Hornik:2008} or CTree
\citep{Hothorn+Hornik+Zeileis:2006}. By applying the presented distributional
trees, a full distributional model is specified in each node of the tree for a
specific distribution family chosen in advance. Hence, selecting the von Mises
distribution allows for an application of distributional trees to circular data
following the tree building algorithm as introduced in Section~2.2. of
\cite{Schlosser+Hothorn+Stauffer:2019}.

In the applied tree-building algorithm, in each node the crucial decision on
how and where to split the data is based on model scores which are obtained by
evaluating the score function $s$ for the individual observations and parameter
estimates (Equation~\ref{eq:scores}). For the von Mises distribution with its
two distribution parameters and a data set of $n$ observations, this yields an
$n \times 2$ matrix that can be employed as a discrepancy measure, capturing
how well each given observation conforms with the estimated location
$\hat{\mu}$ and precision $\hat{\kappa}$, respectively. To capture dependence
on covariates, the association between the model's scores and each available
covariate is assessed using either a parameter instability test~(MOB) or a
permutation test~(CTree). By doing so in each partitioning step, the covariate
with the highest significant association (i.e., lowest significant $p$-value,
if any) is selected for splitting the data. The corresponding split point is
chosen either by optimizing the log-likelihood~(MOB) or a two-sample test
statistic~(CTree) over all possible partitions. This procedure is repeated
recursively until there are no significant parameter instabilities or until
another stopping criterion is met (e.g., subgroup size or tree depth).

@@

\subsection{Data}
\label{sec:wind:data}
The circular response variable considered in this case study is a $10$\,min
mean of wind direction measurements at the airports Innsbruck and Vienna on an
hourly temporal resolution. As predictor variables, we use $1$-hourly resolved
$10$\,min mean observations of various meteorological quantities such as wind
direction, wind speed, temperature, air pressure and humidity, all lagged by
one or three hours according to the respective forecasting step. The
meteorological variables are measured either directly at the airports or within
their vicinities. For Innsbruck, measurements at the airport and along the
intersecting valleys are used, whereas, for Vienna, measurements at the airport
and within an area of about XXX\,km$^2$ are used. A topographical overview of
the airports and their surrounding areas with the station sites employed in
this study is provided in Figure~XXX. In addition, we use derived quantities
such as 3-hourly means, minima and maxima, as well as 1-and 3-hourly temporal
changes and spatial differences towards the airport of the respective
quantities. An overview of the employed data sets can be found in
Table~\ref{tab:data}.

The data used in this study consists of five total years from January 2015 to
December 2018. After first eliminating predictor variables with more than 5\%
missing values and then time points with missing observations, the data set
consists of 41979 time points and 260 covariates for Innsbruck, and of 38985
times points and 494 covariates for Vienna, respectively.

%%
\begin{table}[t!]
  \caption[Table caption text]{Overview of the data sets employed in the case
    study: For Innsbruck and Vienna, various meteorological variables and derived
    quantities of these are considered at the respective stations, located either
    directly at the airports or in their vicinities.}
  \label{tab:data}
  \begin{center}
    \begin{tabular}{l  l}
      \hline
      Data components               & Description \\
      \hline
      Meteorological variables:     & Wind direction, wind (gust) speed, \\
                                    & (reduced) air pressure, relative humidity, \\
                                    & temperature\\ 
      \noalign{\vskip 1mm}                           
      Derived quantities:           & 3-hourly means/minima/maxima, \\
                                    & 1-hourly and 3-hourly temporal changes, \\
                                    & spatial differences towards the airport\\
      \noalign{\vskip 1mm}                              
      Weather stations (Innsbruck): & 4 stations at the airport, \\
                                    & Igls, Kematen, Kufstein, Landeck, Patscherkofel, \\
                                    & Steinach\\
      \noalign{\vskip 1mm}
      Weather stations (Vienna):    & 9 stations at the airport, \\
                                    & Arsenal, Donaufeld, Exelberg, G\"anserndorf, \\
                                    & Gro{\ss}-Enzersdorf, Gumpoldskirchen, Hohe Warte,\\
                                    & Innere Stadt, Jubil\"aumswarte, Mariabrunn, \\
                                    & Seibersdorf, Unterlaa, Wolkersdorf\\
      \hline
    \end{tabular}
  \end{center}
\end{table}
%%

@@

    wide range of applications and climatologies are often also used as a baseline for the

@@

    Section~\ref{sec:introduction}, circular regression models often experience the

@@

    appropriate link function~$g$ in order to guarantee response values within an
    interval of length $2\pi$. As the implementation cannot handle circular covariates,

@@

  \item \emph{Circular distributional forest:} Following the description in

@@

\subsection{Results} 
This section provides a detailed analysis on the predictive performance of the
different proposed statistical models applied to probabilistic wind direction
forecasting. To ensure a comprehensive comparison of the models, wind direction
forecasts are evaluated for two different lead times at two airports with
different climatological site characteristics.
Figure~\ref{fig:boxplot_crpsraw} shows the CRPS values of the employed models
at forecast steps 1\,h and 3\,h for the airports Innsbruck (Panels~a,\,c) and
Vienna (Panels~b,\,d). The scores are aggregated over the five validation years,
yielding yearly mean values for every hour per calendar month, with a lower
score indicating better performance. For both stations at both forecast steps,
the circular forest overall provides the best predictive performance, followed
by the circular distributional tree and, except for the $3$\,h forecast at
Innsbruck, the persistency model. In comparison to the circular tree and
forest, for both stations and forecast steps, the climatological model and the
linear model show clearly higher CRPS values and hence a lower predictive
performance.
%%
\begin{figure}[t]
  \centering
  \setkeys{Gin}{width=1.0\textwidth}
  \includegraphics{_plot_circforest_validation_crpsraw_agg_comparison_with_lowff_v14b.pdf}
  \caption{CRPS skill scores for wind direction forecasts based on the full
    predictive von~Mises distribution for $+1$ and $+3$\,h forecasts at the
    airports Innsbruck and Vienna. Each box-and-whisker contains $24$~hourly scores
    for each of the $12$~months averaged over the $5$~validation years which
    yields a total of $288$~yearly mean values. The scores are shown for the
    climatological, persistency and the linear model as well as for the circular
    distributional tree and forest.}
  \label{fig:boxplot_crpsraw} 
\end{figure}
%%
The different site characteristics of the airports Innsbruck
(Figure~\ref{fig:boxplot_crpsraw}\,a,\,c) and Vienna
(Figure~\ref{fig:boxplot_crpsraw}\,b,\,d) seem to have an effect on the
absolute level of the model performances and on their respective predictive
performance variances. At Innsbruck, due to the surrounding mountains a
limited number of possible wind directions exists, namely the three wind
regimes discussed for Figure~\ref{fig:tree_ibk} in
Section~\ref{sec:introduction}. Therefore, for Innsbruck the wind direction
remains rather constant in one of these possible states, but once a change
takes place it is mostly a major wind direction shift. Due to the few wind
regimes the rather inflexible climatological and linear models score relatively
well with similar CRPS values as the other models
(Figure~\ref{fig:boxplot_crpsraw}\,a,\,c). In addition, at Innsbruck the
potential high prediction errors in case of a paradigm shift seem to lead to a
higher variation in the predictive performance for all models in comparison to
Vienna; this variation is especially high for the persistency model due to its
strong vulnerability to abrupt wind shifts. On the contrary, at Vienna the
geographic surroundings do not constrain any wind direction which leads to
smaller and less abrupt changes in the wind direction as well as less
pronounced wind regimes. This seems to weaken the predictive performance of the
climatological and linear models, and to reduce the performance variability for
all models (Figure~\ref{fig:boxplot_crpsraw}\,b,\,d).

The different forecast steps have apparently only a minor effect on the
predictive performance of the climatological model and the linear model at both
stations. However, for the persistency model, at both stations, higher scores
for the $3$\,h forecast (Figure~\ref{fig:boxplot_crpsraw}\,c,\,d) reveal a
lower performance for longer lead times; this is probably due to the lower 
information content of minimum 3-hourly instead of 1-hourly lagged response 
values employed as covariates in the persistency model. 
The circular distributional tree and forest seem to
partially compensate for the lower skill of the lagged response values by other
covariates, hence their predictive performance only slightly decreases for the
longer lead time. This compensation is especially evident for Innsbruck, where
the performance difference between the persistency model and the tree-based
methods significantly increases from the 1-hourly to the 3-hourly forecast.

In addition to the raw CRPS (Figure~\ref{fig:boxplot_crpsraw}), CRPS skill
scores with the climatological model as a reference are provided in
Figure~\ref{fig:boxplot_crpsskill}. Skill scores are in percent, and positive
values indicate an improvement in the predictive performance over the
reference. For all setups, the circular forest has the highest skill scores
with a mean performance gain of 13--25\% and 58--71\% for Innsbruck and Vienna,
respectively. As discussed for Figure~\ref{fig:boxplot_crpsraw}, this
improvement over the climatological model is lower for Innsbruck due to the low
number of predominant wind regimes and hence a relatively good performance of
the climatological model. Additionally, Figure~\ref{fig:boxplot_crpsskill}
shows that while the persistency model's performance is lower than the
reference (Panel~c) the tree-based models can compensate for the low skill of the 
lagged response values employed as covariates and, hence, are still significantly 
superior to the reference.

%%
\begin{figure}[t]
  \centering
  \setkeys{Gin}{width=1.0\textwidth}
  \includegraphics{_plot_circforest_validation_crpsskill_agg_comparison_with_lowff_v14b.pdf}
  \caption{As Figure~\ref{fig:boxplot_crpsraw}, but showing CRPS skill scores
    with the climatology model as reference. Skill scores are in percent; positive
    values indicate improvements over the reference.} 
  \label{fig:boxplot_crpsskill} 
\end{figure}
%%

% -------------------------------------------------------------------
% Summary
% -------------------------------------------------------------------
\section{Summary and conclusion}
\label{sec:summary}
Circular distributional trees for circular responses are established by
coupling model-based recursive partitioning with the von Mises distribution.
The resulting trees can capture non-linear changes, shifts, and potential
interactions in covariates without prespecification of such effects. This is
particularly useful for modeling a highly fluctuating response such as wind
direction. \dots

@@

\section*{Computational details}
The corresponding implementation of the proposed methodology for circular
trees and forests is provided in the \textsf{R}
package \textbf{circtree} (version~\Sexpr{packageVersion("circtree")}).
% The proposed circular tree and forest models are implemented in the \textsf{R}
% package \textbf{circtree} (version~\Sexpr{packageVersion("circtree")}). 
The package is based on the \textbf{disttree} package
(version~\Sexpr{packageVersion("disttree")}) which applies the main tree
building functions from the \textbf{partykit} package
(version~\Sexpr{packageVersion("partykit")}). All three packages are available
on \textsf{R}-Forge at \url{https://R-Forge.R-project.org/projects/partykit/}.
